---
title: "Getting started with the Marine Sensitivity Atlas"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

The **`msens`** package reads the published **Marine Sensitivity Atlas** (`v8` "marine-atlas") —
species distribution models from many sources merged onto a global 0.05° grid, scored for marine
sensitivity over the US study area, and released as partitioned Parquet on S3 with a
[STAC](https://stacspec.org) catalog and [titiler](https://developmentseed.org/titiler/) tile
services. Everything below runs against the **live public release** — no download, no credentials.

```{r setup}
library(msens)
library(dplyr)
```

## 1. Attach the atlas

A single `attach_atlas()` call configures a DuckDB connection to the S3 release and creates the
table **views**, so every reader and score helper composes off one connection. Use `anon = TRUE`
for anonymous public reads of the released tables.

```{r attach}
con <- attach_atlas(anon = TRUE)

# the atlas tables are now queryable by name
DBI::dbGetQuery(con, "SELECT table_name FROM information_schema.tables
                      WHERE table_schema = 'main' ORDER BY 1")$table_name
```

## 2. Browse — datasets and taxa

```{r datasets}
tbl(con, "dataset") |>
  arrange(sort_order) |>
  select(ds_key, name_short, response_type, native_format) |>
  collect() |>
  knitr::kable()
```

How many merged taxa are valid globally vs. specifically in US waters, per scoring category?

```{r taxa}
tbl(con, "taxon") |>
  group_by(sp_cat) |>
  summarise(
    valid_global = sum(as.integer(is_valid_global), na.rm = TRUE),
    valid_us     = sum(as.integer(is_valid_global & is_valid_usa), na.rm = TRUE),
    .groups = "drop") |>
  arrange(desc(valid_global)) |>
  collect() |>
  knitr::kable()
```

## 3. Search the STAC catalog

Each dataset is a STAC collection; each model is an item whose assets point at the published
Cloud-Optimized GeoTIFFs, PMTiles, and Parquet. Read the **static** catalog with
`rstac::read_stac()` (use `rstac::stac()` only for STAC *API* servers).

```{r stac-r}
library(rstac)
v8 <- read_stac("https://file.marinesensitivity.org/stac/v8/collection.json")

# dataset collections under v8
basename(dirname(vapply(Filter(\(l) l$rel == "child", v8$links), \(l) l$href, character(1))))
```

```{r stac-item}
# one dataset's item + its published assets (native vs. gridded-model representations).
# item links are relative to the collection, so resolve against its base URL.
ch_base   <- "https://file.marinesensitivity.org/stac/v8/ch_fws/"
ch        <- read_stac(paste0(ch_base, "collection.json"))
item_href <- Filter(\(l) l$rel == "item", ch$links)[[1]]$href     # "./…json"
it        <- read_stac(paste0(ch_base, sub("^\\./", "", item_href)))
tibble(
  asset = names(it$assets),
  type  = vapply(it$assets, \(a) if (is.null(a$type)) "" else a$type, character(1)),
  href  = vapply(it$assets, \(a) basename(a$href), character(1))) |>
  knitr::kable()
```

> **Python** reads the same catalog with [`pystac`](https://pystac.readthedocs.io):
>
> ```python
> import pystac
> v8   = pystac.Collection.from_file("https://file.marinesensitivity.org/stac/v8/collection.json")
> item = next(pystac.Collection.from_file(
>     "https://file.marinesensitivity.org/stac/v8/ch_fws/collection.json").get_items())
> {k: a.media_type for k, a in item.assets.items()}
> ```

## 4. Retrieve and map a distribution

The `native_asset` table maps each model to its published tile services. Retrieve a merged taxon's
whole-range Cloud-Optimized GeoTIFF and hand it to `cog_tile_url()`, which builds an XYZ tile
template that drops straight into a leaflet map. (For the US-scored surface instead, `cell_tile_url(mdl_key = mk)` serves the `model_cell` cells.)

```{r map, eval = requireNamespace("leaflet", quietly = TRUE)}
library(leaflet)
mk  <- "ms_merge|WORMS:137209"                    # Dermochelys coriacea (Leatherback Turtle)

# the published whole-range COG for this merged taxon
cog <- tbl(con, "native_asset") |>
  filter(ms_merge_key == mk, ds_key == "ms_merge") |>
  pull(asset_url) |> head(1)
url <- cog_tile_url(cog, colormap = "spectral_r", rescale = c(1, 100))

leaflet() |>
  addProviderTiles("CartoDB.DarkMatter") |>
  addTiles(urlTemplate = url, options = tileOptions(opacity = 0.9)) |>
  setView(0, 20, zoom = 1)
```

## 5. Map the sensitivity score surfaces

Scoring produces a per-cell **value for each metric** — the equal-weight composite marine-sensitivity
score and its per-category components (bird, mammal, fish, …, ecoregion-rescaled to 0–100). These live
in the `cell_metric` table and are served as raster tiles by titiler-v8, which runs a
`cell_id → value` SQL query over the scored cells on demand (there is no pre-rendered COG per metric —
the cells *are* the raster).

```{r score-metrics}
library(glue)

# the composite plus each per-category, ecoregion-rescaled component
score_metrics <- tbl(con, "metric") |>
  filter(grepl("equalweights$|^extrisk_[a-z]+_ecoregion_rescaled$", metric_key)) |>
  pull(metric_key) |> sort()

# a readable label from the metric_key (descriptions are sparse in the release)
label_of <- function(mk) ifelse(
  grepl("equalweights$", mk), "composite (equal-weight)",
  paste0(sub("^extrisk_([a-z]+)_.*$", "\\1", mk), " extinction-risk"))

# the (cell_id, value) SELECT titiler renders — the same path the scores app uses
cell_sql <- function(mk) glue(
  "SELECT cm.cell_id, cm.val AS value FROM cell_metric cm ",
  "JOIN metric m ON cm.metric_seq = m.metric_seq WHERE m.metric_key = '{mk}'")
```

Ask titiler for each surface's **raster statistics** — proof the served tiles carry real per-cell
values (not a static image), computed live over the scored cells:

```{r score-stats}
stats <- lapply(score_metrics, \(mk) cell_stats(cell_sql(mk)))
names(stats) <- score_metrics

purrr::imap_dfr(stats, \(s, mk) tibble::tibble(
  surface = label_of(mk), cells = s$n,
  min = round(s$min, 1), mean = round(s$mean, 1),
  median = round(s$p50, 1), p98 = round(s$p98, 1), max = round(s$max, 1))) |>
  arrange(desc(mean)) |>
  knitr::kable(caption = "titiler raster statistics per score surface (US study area)")
```

Hand each metric's SQL to `cell_tile_url()` for a leaflet layer, rescaled to its own range. A layers
control toggles between the composite and its components — one raster surface at a time:

```{r score-map, eval = requireNamespace("leaflet", quietly = TRUE)}
# add components first, composite last (so it is the default-visible base layer)
ordered <- c(setdiff(score_metrics, grep("equalweights$", score_metrics, value = TRUE)),
             grep("equalweights$", score_metrics, value = TRUE))

m <- leaflet() |> addProviderTiles("CartoDB.DarkMatter") |> setView(-105, 38, zoom = 3)
for (mk in ordered) {
  s <- stats[[mk]]
  m <- m |> addTiles(
    urlTemplate = cell_tile_url(cell_sql(mk), colormap = "spectral_r",
                                rescale = c(s$min, s$max)),
    group = label_of(mk), options = tileOptions(opacity = 0.9))
}
m |> addLayersControl(
  baseGroups = label_of(rev(ordered)),                 # composite listed first
  options = layersControlOptions(collapsed = FALSE))
```

## 6. Score a Program Area

`scores_for_pra()` returns the per-category marine-sensitivity scores for a BOEM Program Area —
ready for a flower plot, where each petal's height is the category score.

```{r scores}
sc <- scores_for_pra(con, "ALA")                  # ALA = a BOEM Program Area key
knitr::kable(sc[c("component", "score")], digits = 1)
```

```{r flower, eval = requireNamespace("ggiraph", quietly = TRUE), fig.height = 4.5}
plot_flower(
  sc,
  fld_category = component,
  fld_height   = score,
  fld_width    = even,
  tooltip_expr = "{component}: {round(score, 1)}",
  title        = "ALA")
```

## 7. Species

The `taxon` table carries each merged taxon's category, governing extinction-risk score, and
protection flags — e.g. the highest-risk mammals in US waters:

```{r species}
tbl(con, "taxon") |>
  filter(sp_cat == "mammal", is_valid_usa, er_score >= 80) |>
  select(scientific_name, common_name, extrisk_code, er_score, is_mmpa) |>
  arrange(desc(er_score), scientific_name) |>
  head(10) |>
  collect() |>
  knitr::kable()
```

To go the other way — *which species occur in a set of cells* — use `species_for_cells(con, cells)`,
which joins the merged `model_cell` surface to `taxon` (that view needs credentials, so call
`attach_atlas()` without `anon`).

```{r cleanup, include = FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
